# `r params$dataset_name`

```{r}
#| label: load_libraries
#| message: false
#| warning: false
#| echo: false

library(knitr)
library(kableExtra)

pacman::p_load(readr, dplyr, ggplot2,
               glmmTMB, modelbased, see, 
               sdmTMB, sf, leaflet, glue)

theme_set(theme_bw(base_size = 13) #+
            #theme(legend.position="bottom")
            )
```

```{r}
#| label: load-data
#| message: false
#| echo: false


# Anchors the project root explicitly to nw_atlantic_kelp/, since this file
# runs (via {{< include >}}) with its working directory set to
# markdown/models_and_data/ — plain here::here() would otherwise resolve
# root to that folder instead of climbing up to the real project root.
# If this file moves, update the path below to match.
# advice from Claude after hours of stackoverflow and other googling fails
here::i_am("markdown/models_and_data/_analysis-template.qmd")

source(here::here("scripts/load_nwa_data.R"))
#df <- read.csv(here::here(params$path))


kelptime_study <- nwa_dat |>
  dplyr::filter(study == params$dataset_name)

study_data <- read_csv(here::here(params$path),
  show_col_types = FALSE
 ) |>
  filter(Study == params$dataset_name) |> # needed for mined data
  dplyr::select(-Sample.ID) |> # don't need and was causing problems
  mutate(
         Site = iconv(Site,  from = "latin1", to = "UTF-8", sub = ""),
         Site = gsub("\\x89", "", Site ),
         Site = gsub("\\n", "", Site ),
         Site = stringi::stri_enc_toutf8(Site))

```


## The Data

```{r}
#| label: is this mined?
#| output: asis
#| echo: false
#| 
if(grepl("mined", params$path)){
  cat("**NOTE**: This data was mined from papers, and was not given to us by the author.")
}
```


Here is your dataset as reprocessed  Please take a look.

::: {.callout-note collapse="true"}

### Click to see the clean data
```{r}
#| label: show_study_data
# 
 study_data |>
          kable( format = "html") %>%
          kable_styling(position = "center", full_width = F, bootstrap_options = c("striped","hover"))%>%
          scroll_box( width = "1000px", height = "500px", fixed_thead = T)

```
:::

Here is your dataset as where we have now summed across all kelp species and put into the KelpTime main database format.

::: {.callout-note collapse="true"}

### Click to see the data in kelptime format
```{r}
#| label: show_kelptime_data
#|
kelptime_study |>
         kable( format = "html") %>% 
         kable_styling(position = "center", full_width = F, bootstrap_options = c("striped","hover"))%>%        
         scroll_box( width = "1000px", height = "500px", fixed_thead = T)
```
:::

## Map of Sites

Here are the sites from `r params$dataset_name`

```{r}
#| label: site_map
#| 
# make sf obj
kelptime_study <- kelptime_study |> st_as_sf(
  coords = c("longitude", "latitude"),
  crs = 4326
)

# colors!
good_cols <- colors() |> 
  stringr::str_subset("gr[ea]y", negate = TRUE) |>
  stringr::str_subset("white", negate = TRUE) |>
  stringr::str_subset("black", negate = TRUE) |>
  gplots::col2hex()

# get site-level LL
kelptime_study_grouped <- kelptime_study |>
  group_by(trajectory) |>
  summarize(study = study[1],
            site = site[1],
            min_year = min(sample_year),
            max_year = max(sample_year),
            n = n(),
            eco_collapsed = eco_collapsed[1]) |>
  mutate(text = glue("{trajectory}<br>study: {study}<br>{min_year} - {max_year}, n = {n}<br>site: {site}<br>{eco_collapsed}"),
         col = good_cols[rank(trajectory) |> round()])

leaflet() |>
  addTiles() |>
  addCircleMarkers(data = kelptime_study_grouped |> st_centroid(),
             fillColor = ~col,
             fillOpacity = 0.8,
             popup = ~text,
             stroke = 0)
```

## Plots of Raw Data

## How Does the Data Fit Into the Long-Term Trend Analysis?

We fit two models. One where the slope varies by ecoregion (note: the GOSL and Newfoundland are collapsed) and one where the slope does not vary by ecoregion. Based on the modeled results, here is the smoothed fits to your data. Note -  the model with a common slope appears to best predict (best fit in terms of prediction and p-value for slope*ecoregion is not distinguishable from 0).

**Note** - the study (black) line might not line up in the middle or on trajectories, as it is shrunken in towards the grand mean for the ecoregion.

### 3.1 Slope Varying by Ecoregion Model

```{r}
#| label: plot_fit_vary
#| fig-width: 6

mod_ecoregion <- readRDS(here::here("models/mod_ecoregion.rds"))


# for plotting to do the back transform
exp_trans <- function(dat){

  dat |>
    mutate(
      Predicted = exp(Predicted)-0.01,
           CI_low = exp(CI_low)-0.01,
           CI_high = exp(CI_high)-0.01,
           CI_high = ifelse(CI_high>1.1, 1.1, CI_high)
    )|>
               mutate(trajectory = 
                        gsub(study[1], "", trajectory))

}


re_ecoregion <- ranef(mod_ecoregion)$cond$study |>
  as_tibble(rownames = "study")  |> 
  filter(study == params$dataset_name) |> 
  pull(`(Intercept)`)


dat_fixed_ecoregion <- estimate_relation(mod_ecoregion,
                               include_random = FALSE,
                  length = 200,
                  data = kelptime_study) |>
  mutate(Predicted = Predicted + re_ecoregion,
         CI_high = CI_high + re_ecoregion,
         CI_low = CI_low + re_ecoregion) |>
  exp_trans()

dat_traj_ecoregion <- estimate_relation(mod_ecoregion,
                              include_random = TRUE,
                  length = 200,
                  data = kelptime_study) |>
    exp_trans()


ggplot(dat_traj_ecoregion, aes(x = year)) +
  geom_line(aes(y = Predicted,
                color = trajectory), alpha = 0.5) +
  geom_point(aes(y = focal_std_by_ecoregion,
                 group = trajectory,
                 color = trajectory), alpha = 0.5,
             data = kelptime_study |>
               mutate(trajectory = 
                        gsub(study[1], "", trajectory))) +
  geom_line(data = dat_fixed_ecoregion,
            aes(y = Predicted), color = "black") +
  geom_ribbon(data = dat_fixed_ecoregion,
            aes(y = Predicted,
                ymin = CI_low,
                ymax = CI_high), color = NA,
            alpha = 0.4) +
  labs(x = "", y = "Standardized Kelp",
      color = "site") +
  guides(color = "none")

```

### 3.2 Common Slope Model
```{r}
#| label: plot_fit_common
#| fig-width: 6

mod_common_slope <- readRDS(here::here("models/mod_common_slope.rds"))

re <- ranef(mod_common_slope)$cond$study |>
  as_tibble(rownames = "study")  |> 
  filter(study == params$dataset_name) |> 
  pull(`(Intercept)`)


dat_fixed <- estimate_relation(mod_common_slope,
                               include_random = FALSE,
                  length = 200,
                  data = kelptime_study) |>
  mutate(Predicted = Predicted + re,
         CI_high = CI_high + re_ecoregion,
         CI_low = CI_low + re_ecoregion) |>
  exp_trans()

dat_traj <- estimate_relation(mod_common_slope,
                              include_random = TRUE,
                  length = 200,
                  data = kelptime_study) |>
    exp_trans()

ggplot(dat_traj, aes(x = year)) +
  geom_line(aes(y = Predicted,
                color = trajectory), alpha = 0.5) +
  geom_point(aes(y = focal_std_by_ecoregion,
                 group = trajectory,
                 color = trajectory), alpha = 0.5,
             data = kelptime_study |>
               mutate(trajectory = 
                        gsub(study[1], "", trajectory))) +
  geom_line(data = dat_fixed,
            aes(y = Predicted), color = "black") +
  geom_ribbon(data = dat_fixed,
            aes(y = Predicted,
                ymin = CI_low,
                ymax = CI_high), color = NA,
            alpha = 0.4) +
  labs(x = "", y = "Standardized Kelp",
      color = "site") +
  guides(color = "none")
```

## How Does the Data Fit Into the Spatially Varying Trend Analysis?

First, let's look at the map showing slopes for each site.

```{r}
#| label: load_trends_spatial_fit

fitted_spatial_predictions <- readRDS(here::here("data/model_derived_data/fitted_spatial_model_sf.rds")) |>
  filter(study == params$dataset_name)

slopes_spatial_fits <- 
  fitted_spatial_predictions |>
  group_by(trajectory) |>
  slice(1L) |>
  ungroup() |>
  select(trajectory, latitude, eco_collapsed, combined_slope_median:combined_slope_upr) |>
  arrange(combined_slope_mean) |>
  mutate(samp = 1:n()) |>
  st_transform(4326) |>
  mutate(text = glue("slope = {round(combined_slope_mean,4)}"))

```

``` {r}
#| label: plot_spatial_trends

domain <- c(-1*max(abs(slopes_spatial_fits$combined_slope_mean)),
               max(abs(slopes_spatial_fits$combined_slope_mean)))

spatial_slope_pal <- 
  colorNumeric(
    colorspace::divergingx_hcl(7,
                               palette = "RdYlBu"),
    domain = domain)

leaflet() |>
  addTiles() |>
  addCircleMarkers(data = slopes_spatial_fits,
             fillColor = ~spatial_slope_pal(combined_slope_mean),
             fillOpacity = 0.8,
             popup = ~text,
             stroke = 0) |>
   addLegend(pal =spatial_slope_pal,
             values = seq(domain[1], domain[2], length.out = 7) )
```

Next, let's see how that actually plays out on a scatter plot with individual fit lines overlain. Note, colors go from south to north.

```{r}
#| label: fit_spatial_curve_plot

ggplot(fitted_spatial_predictions,
         aes(x = year,
             y = exp(est)-0.01,
             color = latitude,
             group = trajectory,
         )) +
    geom_line(linewidth = 1.1, alpha = 0.9) +
    labs(y = "Standardized Kelp Abundance", x = "",
         color = "Latitude") +
    theme_bw(base_size = 14)+
    scale_colour_distiller(palette = "RdBu", direction = 1) +
    geom_point(aes(y = yi), alpha = 0.4) +
    facet_wrap(vars(eco_collapsed))

```

The above shows the results directly from the fit model for the individual data sets. These model, however, has a random effect for each site. Let's instead show a map of the region with the general trends from the region from the model fit.

Note - circle markers with an outline around them are estimates from the model for observed sites. Other points are from a grid around the region and represent model estimates for those sites.

```{r}
#| label: load_trends_spatial_predict

spatial_predictions <- readRDS(here::here("data/model_derived_data/prediction_spatial_model_sf.rds")) 

# crop to extent of dataset
spatial_predictions <- spatial_predictions |> 
  st_crop(st_bbox(fitted_spatial_predictions) |>st_as_sfc() |> st_buffer(1e1))

slopes_spatial_predictions <- 
  spatial_predictions |>
  group_by(X, Y) |>
  slice(1L) |>
  ungroup() |>
  select(X, Y, combined_slope_median:combined_slope_upr) |>
  arrange(combined_slope_mean) |>
  mutate(samp = 1:n()) |>
  st_transform(4326) |>
  mutate(text = glue("slope = {round(combined_slope_mean,4)}"))

```

``` {r}
#| label: plot_spatial_predictions

domain <- c(-1*max(abs(slopes_spatial_predictions$combined_slope_mean)),
               max(abs(slopes_spatial_predictions$combined_slope_mean)))

spatial_slope_pal <- 
  colorNumeric(
    colorspace::divergingx_hcl(7,
                               palette = "RdYlBu"),
    domain = domain)

leaflet() |>
  addTiles() |>
  addCircleMarkers(data = slopes_spatial_predictions,
             fillColor = ~spatial_slope_pal(combined_slope_mean),
             fillOpacity = 0.8,
             popup = ~text,
             stroke = 0) |>
    addCircleMarkers(data = slopes_spatial_fits,
             fillColor = ~spatial_slope_pal(combined_slope_mean),
             fillOpacity = 0.8,
             popup = ~text,
             stroke = 1) |>
   addLegend(pal =spatial_slope_pal,
             values = seq(domain[1], domain[2], length.out = 7) )
```


Last, let's show the data set, but the curves show the regional fits - again, colors go south to north.

```{r}
#| label: predicted_spatial_curve_plot
#| 
spatial_curves <- readRDS(here::here("data/model_derived_data/prediction_spatial_model_curves.rds")) |>
  st_as_sf(coords = c("X", "Y"), 
           crs = st_crs(spatial_predictions),
           remove = FALSE)|> 
  # crop to the dataset
  st_crop(st_bbox(fitted_spatial_predictions) |>st_as_sfc() |> st_buffer(1e1)) |>
  st_transform(4326) %>%
    mutate(lon = sf::st_coordinates(.)[,1],
                lat = sf::st_coordinates(.)[,2]) |>
  # Trim to the time of the data set
  filter(year >= min(study_data$Sample.Year)) |>
  filter(year <= max(study_data$Sample.Year))

ggplot(fitted_spatial_predictions,
       aes(x = year, color = latitude,
                    y = yi)) +
  geom_point( alpha = 0.9) +
  geom_line(data = spatial_curves,
            aes(y = est, 
                color = lat,
                group = Y))  +
  scale_color_viridis_c() +
  labs(x="", y = "Standardized Kelp Abundance", 
       color = "Latitude")

```

## How Does the Data Fit Into the Spatio-Temporal Time Series Model?


First, let's look at the map showing slopes for each site for each decade.

```{r}
#| label: splot_by_decade_fit

yrs <- c(1975, 1985, 1995, 2005, 2015) + 5
```

Next, let's see how that actually plays out on a scatter plot with individual fit lines overlain. Note, colors go from south to north.

The above shows the results directly from the fit model for the individual data sets. These data, however, have a random effect for each site. Let's instead show a map of the region by decade with the general trends from the region from the model fit.

Last, let's show the data set, but the curves show the regional fits - again, colors go south to north.